---
title: "Bayesian Linear Models"
subtitle: "02 - BRM basics"
author: "Stefano Coretta"
institute: "University of Edinburgh"
date: "2023/05/16"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - default
      - "../custom.css"
      - "../xaringan-themer.css"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "../macros.js"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, out.height = "500px", dpi = 300,
  fig.align = "center", fig.width = 7, fig.height = 5
)
knitr::opts_knit$set(root.dir = here::here())

library(tidyverse)
theme_set(theme_minimal())
library(glue)

library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))

options(htmltools.dir.version = FALSE)

library(brms)

my_seed <- 9899
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i", "700", "700i"),
  code_font_google   = google_font("Fira Mono"),
  outfile = "./slides/xaringan-themer.css"
)
```

```{r mald, echo=FALSE}
mald <- readRDS("./data/mald.rds")
```

```{r rt-dist-2}
rt_mean <- mean(mald$RT)
rt_sd <- sd(mald$RT)
rt_mean_text <- glue("mean: {round(rt_mean)} ms")
rt_sd_text <- glue("SD: {round(rt_sd)} ms")

rt_dist_2 <- ggplot(data = tibble(x = -500:2500), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(rt_mean, rt_sd), colour = "#9970ab", linewidth = 1.5) +
  scale_x_continuous(n.breaks = 5) +
  geom_vline(xintercept = rt_mean, colour = "#1b7837", linewidth = 1) +
  annotate(
    "label", x = rt_mean + 1, y = 0.0006,
    label = rt_mean_text,
    fill = "#1b7837", colour = "white"
  ) +
  annotate(
    "label", x = 2000, y = 0.0006,
    label = rt_sd_text,
    fill = "#8c510a", colour = "white"
  ) +
  annotate(
    "label", x = 5, y = 0.0006,
    label = "theoretical sample\ndistribution",
    fill = "#9970ab", colour = "white"
  ) +
  labs(
    title = "Theoretical sample distribution of RT",
    subtitle = glue("Gaussian distribution: mean = {round(rt_mean)} ms, SD = {round(rt_sd)}"),
    x = "RT (ms)", y = "Relative probability"
  )

rt_dist_2 +
  geom_density(data = mald, aes(RT), colour = "grey", fill = "grey", alpha = 0.2) +
  geom_rug(data = mald, aes(RT), alpha = 0.05) +
  annotate(
    "label", x = 20, y = 0.0015,
    label = "empirical sample\ndistribution",
    fill = "grey", colour = "white"
  )
```


---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & = 1046 \\
\sigma & = 348
\end{aligned}
$$
]

--

- The RT values are distributed according to a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$.

- The mean and SD from the sample do not take into consideration the **uncertainty and variability** in the sample.

---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & \sim Gaussian(\mu_1, \sigma_1) \\
\sigma & \sim Gaussian_+(\mu_1, \sigma_1)
\end{aligned}
$$
]

--

- We can account for uncertainty and variability by assuming that the mean and SD are themselves values coming from a probability distribution.

- The mean $\mu$ is a value from a Gaussian distribution with mean $\mu_1$ and SD $\sigma_1$.

- The SD $\sigma$ is a value from a Gaussian distribution with mean $\mu_2$ and SD $\sigma_2$.

  - SD can only be positive so the Gaussian distribution is truncated to positive values only ($Gaussian_+$).

---

```{r rt-bm, echo=FALSE}
rt_bm <- brm(
  RT ~ 1,
  data = mald,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/rt_bm",
  seed = my_seed
)
```

```{r rt-bm-code, eval=FALSE, echo=TRUE}
# Attach the brms package
library(brms)

# Run a Bayesian model
rt_bm <- brm(
  # This is the formula of the model.
  RT ~ 1,
  # This is the probability distribution family.
  family = gaussian(),
  # And the data.
  data = mald
)
```
