---
title: "02 - Regression"
author: "Stefano Coretta"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_light())
library(brms)
```

## Vowel duration

```{r}
#| label: durations
#| echo: false
#| message: false

durations <- read_csv("data/coretta2018/token-measures.csv") |> 
  filter(language == "Italian")

durations
```

## Vowel durations: plot

```{r}
#| label: fig-durations
#| fig-cap: "Vowel duration and speech rate."
#| warning: false
#| echo: false

durations |> 
  ggplot(aes(speech_rate, v1_duration)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Speech rate (syl/s)", y = "Vowel duration (ms)"
  )
```

## A log-normal model of vowel duration

$$
dur \sim LogNormal(\mu, \sigma)
$$

. . . 

```{r}
#| label: dur-logn
#| eval: false

dur_logn <- brm(
  v1_duration ~ 1,
  family = lognormal,
  data = durations
)
```

. . .

::: box-error
But we want to investigate the relationship between speech rate and vowel duration!
:::

## Allow $\mu$ to vary depending on speech rate

$$
\begin{align}
dur_i & \sim LogNormal(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot SR_i\\
\end{align}
$$

. . .

::: box-tip
Does the formula for $\mu$ ring a bell?
:::

## The formula of a line

::: box-note
$$
y = a + b \cdot x
$$

- $a$ is the line's **intercept**. This is $y$ when $x$ is 0.

- $b$ is the line's **slope** (aka gradient). This is the change in $y$ for every unit increase of $x$.

See [Linear models illustrated](https://stefanocoretta.shinyapps.io/lines/).

:::

## Regression model of vowel duration

$$
\begin{align}
dur_i & \sim LogNormal(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot SR_i\\
\end{align}
$$

::: box-note
- $\beta_0$ is the intercept. This is the mean vowel duration when speech rate is 0.

- $\beta_1$ is the slope. This is the change in vowel duration for each unit increase of speech rate (syl/s).
:::

## But...

::: box-warning
**Speech rate 0 doesn't make sense!**

- Speech rate cannot be zero syllables per second.
:::

. . .

::: box-tip
We can **centre** speech rate.

- Subtract the mean speech rate from all the speech rate values.
:::

## Centred speech rate

```{r}
#| label: centre

mean(durations$speech_rate, na.rm = TRUE)

durations <- durations |> 
  mutate(
    speech_rate_c = speech_rate - mean(speech_rate, na.rm = TRUE)
  )
```

::: box-note
- `speach_rate_c` = 0 means mean speech rate. 
:::

## Centred speech rate: plot

```{r}
#| label: fig-durations-centred
#| fig-cap: "Vowel duration and (centred) speech rate."
#| warning: false
#| echo: false

durations |> 
  ggplot(aes(speech_rate_c, v1_duration)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Speech rate (syl/s centred)", y = "Vowel duration (ms)"
  )
```

## Regression model of vowel durations: centred speech rate

$$
\begin{align}
dur_i & \sim LogNormal(\mu_i, \sigma)\\
\mu_i & = \beta_0 + \beta_1 \cdot (SR_{ctr[i]})\\
\end{align}
$$

::: box-note
- $\beta_0$ is the intercept. This is the mean RT when centred speech rate is 0 (i.e. when speech rate is at its mean = `{r} mean(durations$speech_rate)`).

- $\beta_1$ is the slope. This is the change in RT for each unit increase of centred speech rate (i.e for every unit increase of speech rate).
:::

## Regression model of vowel durations: code

```{r}
#| label: rt-phonlev

dur_sr <- brm(
  v1_duration ~ 1 + speech_rate_c,
  family = lognormal,
  data = durations,
  cores = 4,
  seed = 1032,
  file = "data/cache/dur_sr"
)
```

```{r}
#| label: dur-sr-summary

summary(dur_sr)
```

```{r}
#| label: fig-dur-sr-cond

conditional_effects(dur_sr)
```

```{r}
#| label: fig-dur-sr-pp
#| fig-cap: "Posterior predictive check plot of `dur_sr`."
pp_check(dur_sr, ndraws = 20)
```

