---
title: "06 - What's next?"
author: "Stefano Coretta"
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_light())
library(brms)
```

```{r}
#| label: mald
#| include: false

mald <- readRDS("data/tucker2019/mald_1_1.rds")
```

## This is just the beginning

You have learnt a great deal already!

::: box-note
-   Statistics as a tool to deal with uncertainty and variability.

-   Regression models (Gaussian, log-normal, Bernoulli, Poisson, negative binomial, ordinal) for a variety of outcome variables.

-   One predictor: numeric or categorical.

-   Working with posterior draws.
:::

. . .

But there is more you need to learn to use regression models in actual research!

## Many predictors: interactions

```{r}
#| label: fig-mald-int
#| fig-cap: "Regression lines illustrating an interaction between phonetic distance and lexical status."
#| message: false
#| echo: false

mald |> 
  ggplot(aes(PhonLev, RT)) +
  geom_smooth(aes(colour = IsWord), method = "lm") +
  labs(
    x = "Phonetic distance", y = "RT (ms)",
    colour = "Real\nwords"
  )
```

## Other types of outcome variables

::: box-note
-   Variables bounded between 0 and 1, like proportions. Use a beta regression.

-   Categorical variables with more than two categories (but unordered). Use a categorical (multinomial) regression.

-   Multiple outcome variables! Use multivariate regression models.

-   Can you think of more types?
:::

## Non-linear effects: smooth terms

```{r}
#| label: durations
#| include: false

durations <- read_csv("data/coretta2018/token-measures.csv") |> 
  filter(language == "Italian")
```

```{r}
#| label: fig-durations
#| fig-cap: "Vowel duration and speech rate."
#| warning: false
#| echo: false
#| message: false

durations |> 
  ggplot(aes(speech_rate, v1_duration)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", colour = "red") +
  geom_smooth(method = "gam") +
  labs(
    x = "Speech rate (syl/s)", y = "Vowel duration (ms)"
  )
```

## Repeated measures: multilevel/hierarchical/mixed-effects

```{r}
#| label: fig-mald-subjs
#| fig-cap: "Jitter plot showing RTs by lexical status for a selection of participants."
#| message: false

set.seed(3198)
mald |> 
  filter(Subject %in% sample(unique(mald$Subject), 12)) |> 
  ggplot(aes(IsWord, RT)) +
  geom_jitter(alpha = 0.25, width = 0.2) +
  stat_summary(aes(colour = IsWord)) +
  labs(
    x = "Phonetic distance", y = "RT (ms)",
    colour = "Real\nwords"
  ) +
  facet_wrap(vars(Subject))
```

## Model diagnostics

::: box-note
-   Model diagnostics help us determine cases when there is something wrong with the model or the data or both.

-   The posterior predictive check plot is one type of diagnostic, but there are others (like $\hat{R}$ and Effective Sample Size in the model summary).
:::

## Prior probability distributions

::: box-note
-   We've been using the default priors set by brms. They are usually fine.

-   It might be preferable to specify custom priors.

-   In any case, learning the basics of prior specification is fundamental.
:::

## Frequentist vs Bayesian statistics

::: box-note
-   Modern research is dominated by the "null ritual", a degenerate form of frequentist statistics [@gigerenzer2004; @gigerenzer2004a; @gigerenzer2018]. It is based on "rejecting a nil hypothesis".

-   The "null ritual" is **not** a robust statistical approach.

-   Proper frequentist methods are difficult to apply in practice.

-   Bayesian statistics is about estimating uncertainty and accounting for variability with posterior probability distributions.
:::
