---
title: "Tutorial 02"
execute: 
  cache: true
---

```{r}
#| label: pkgs
#| message: false
#| echo: false

library(tidyverse)
theme_set(theme_light())

library(brms)

polite <- read_csv("data/polite.csv") %>%
  drop_na()
```

## Gender and register

The `polite` data also has information on the gender of the participant (`F` for female and `M` for male).

Let's plot again `f0mn` and `attitude` (register), this time including `gender`.

```{r}
#| label: gender-plot
#| echo: true

polite %>%
  ggplot(aes(gender, f0mn, colour = attitude)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.7))
```

The plot suggests that `f0mn` might be somewhat lower in the polite register in female participants, but not really in male participants.

How do we account for this potential **interaction** between register and gender?

## Interactions

When there is the potential for two or more predictors in a model to "interact" with each other, you should include a so-called **interaction**.

Here's how:

```{r}
#| label: f0mn-bm-2
#| eval: false

f0mn_bm_2 <- brm(
  f0mn ~ attitude + gender + attitude:gender,
  family = gaussian(),
  data = polite,
  backend = "cmdstanr"
)
```

```{r}
#| label: f0mn-bm-2-2
#| echo: false
#| message: false

f0mn_bm_2 <- brm(
  f0mn ~ attitude + gender + attitude:gender,
  family = gaussian(),
  data = polite,
  backend = "cmdstanr",
  file = "data/cache/f0mn_bm_2.rds"
)
```

In the model's formula we have `attitude`, `gender` and something new: `attitude:gender`.

The term `attitude:gender` is an interaction: it tells the model to allow `attitude` and `gender` to interact, so that if the effect of one predictor differs depending on the effect of the other, the model will be able to capture this.

In other words, if the effect of `attitude` differs by `gender`, we need to include an interaction in the model to capture this. You normally always include interactions between predictors and the results of the interaction term will help you understand if there is an interaction or not.

Now, this is the model summary.

```{r}
#| label: f0mn-bm-2-summ

summary(f0mn_bm_2)
```

Let's go through the population-level effects:

- `Intercept`: this is the mean `f0mn` when [attitude=informal] and [gender=F]. There is a 95% probability that this is between 247 and 266 hz.

- `attitudepol`: this is the difference in mean `f0mn` when you go from the `Intercept` to [attitude=polite] (so [gender=F]). We can be 95% confident that this difference is between -31 and -4 hz. In other words, when the register is polite, `f0mean` lowers by 4-31 hz in female participants at 95% probability.

- `genderM`: this is the difference in mean `f0mn` when you go from the `Intercept` to [gender=M] (so [attitude=inf]). There is a 95% probability that this difference is between -135 and -105 hz.

- Now, the interaction term `attitudepol:genderM`: this tells you the adjustment you need to make to the effect of `attitudepol` to get the effect when [gender=M].

What does it mean? It means that, on average, the effect of `attitude` is 6.7 hz smaller in males than in females: -17.56 + 6.7 = -10.86. However, there is also the possibility that the effect is 0 or that it is positive rather than negative, because the 95% CrI go from -14 to +27 hz.

## Plotting the model results

We can visualise the results using the `conditional_effects()` function.

```{r}
#| label: f0-bm-2-cond

conditional_effects(f0mn_bm_2, "gender:attitude")
```

The effect of attitude is clearer in female participants than in males, although it is still possible that male participants lower their mean f0 when speaking politely.

## Fomula of an interaction

We can understand what the interaction term is doing through the formula of the model we fitted above. Feel free to skip this part if you find that formulae don't help you.

$$
\begin{aligned}
f0mn & \sim Gaussian(\mu, \sigma) \\
\\
\mu & \sim \beta_0 + \beta_1 \times attitude_{pol} + \beta_2 \times gender_{M} + \beta_3 \times attitude_{pol} \times gender_{M} \\
\sigma & \sim Gaussian_+(39, 1.97)
\end{aligned}
$$

When the [attitude=polite], $attitude_{pol}$ is 1 and when [gender=M], $gender_M$ is 1, otherwise they are 0.

Now, if you want to get $\mu$ (i.e. the mean `f0mn`) for a particular register/gender, you just fill in the estimates from the model summary:

- $\beta_0$ is the `Intercept`.
- $\beta_2$ is `attitudepol`.
- $\beta_3$ is `genderM`.
- $\beta_4$ is `attitudepol:genderM`

Let me show you the process for the mean `f0mn` in the polite register and female participant.

$$
\begin{aligned}
\mu_{[polite, F]} = & \beta_0 + \beta_1 \times attitude_{pol} + \beta_2 \times gender_{M} + \beta_3 \times attitude_{pol} \times gender_{M} \\
 = & 256.72 + (-17.56) \times 1 + (-119.61) \times 0 + 6.7 \times 1 \times 0 \\
 = & 256.72 + (-17.56) \times 1 \\
 = & 239.16
\end{aligned}
$$
So, the mean `f0mn` when the register is polite and the gender is female is about 239 hz.

Try and figure out the mean `f0mn` for the other combinations of register and gender.

## Challenge

If you feel like and you have the time, try running the model again but this time use `musicstudent` (whether the participant studied music or not) and `months_ger` (how many months the participant have spent in Germany).

Remember to include an interaction between the two predictors. A short cut is to write `musicstudent * months_ger` instead of the longer `musicstudent + months_ger + musicstudent:months_ger`.
