---
title: "Bayesian Linear Models"
subtitle: "02 - BRM basics"
author: "Stefano Coretta"
institute: "University of Edinburgh"
date: "2023/07/07"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - default
      - "../custom.css"
      - "../xaringan-themer.css"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "../macros.js"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, out.height = "500px", dpi = 300,
  fig.align = "center", fig.width = 7, fig.height = 5
)
knitr::opts_knit$set(root.dir = here::here())

library(tidyverse)
theme_set(theme_minimal())
library(glue)

library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))

options(htmltools.dir.version = FALSE)

library(brms)

my_seed <- 9899
```

```{r mald, echo=FALSE}
mald <- readRDS("./data/mald.rds")
```

layout: true

## Estimate mean and SD

---

```{r rt-dist-2}
rt_mean <- mean(mald$RT)
rt_sd <- sd(mald$RT)
rt_mean_text <- glue("mean: {round(rt_mean)} ms")
rt_sd_text <- glue("SD: {round(rt_sd)} ms")

rt_dist_2 <- ggplot(data = tibble(x = -500:2500), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(rt_mean, rt_sd), colour = "#9970ab", linewidth = 1.5) +
  scale_x_continuous(n.breaks = 5) +
  geom_vline(xintercept = rt_mean, colour = "#1b7837", linewidth = 1) +
  annotate(
    "label", x = rt_mean + 1, y = 0.0006,
    label = rt_mean_text,
    fill = "#1b7837", colour = "white"
  ) +
  annotate(
    "label", x = 2000, y = 0.0006,
    label = rt_sd_text,
    fill = "#8c510a", colour = "white"
  ) +
  annotate(
    "label", x = 5, y = 0.0006,
    label = "theoretical sample\ndistribution",
    fill = "#9970ab", colour = "white"
  ) +
  labs(
    title = "Theoretical sample distribution of RT",
    subtitle = glue("Gaussian distribution: mean = {round(rt_mean)} ms, SD = {round(rt_sd)}"),
    x = "RT (ms)", y = "Relative probability"
  )

rt_dist_2 +
  geom_density(data = mald, aes(RT), colour = "grey", fill = "grey", alpha = 0.2) +
  geom_rug(data = mald, aes(RT), alpha = 0.05) +
  annotate(
    "label", x = 20, y = 0.0015,
    label = "empirical sample\ndistribution",
    fill = "grey", colour = "white"
  )
```


---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & = 1046 \\
\sigma & = 348
\end{aligned}
$$
]

--

- The RT values are distributed according to a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$.

- The mean and SD from the sample do not take into consideration the **uncertainty and variability** in the sample.

---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & \sim Gaussian(\mu_1, \sigma_1) \\
\sigma & \sim Gaussian_+(\mu_2, \sigma_2)
\end{aligned}
$$
]

--

- We can account for uncertainty and variability by assuming that the mean and SD are themselves values coming from a probability distribution.

- The mean $\mu$ is a value from a Gaussian distribution with mean $\mu_1$ and SD $\sigma_1$.

- The SD $\sigma$ is a value from a Gaussian distribution with mean $\mu_2$ and SD $\sigma_2$.

  - SD can only be positive so the Gaussian distribution is truncated to positive values only ( $Gaussian_+$ ).

---

```{r rt-bm, echo=FALSE}
rt_bm <- brm(
  RT ~ 1,
  data = mald,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/rt_bm",
  seed = my_seed
)
```

```{r rt-bm-code, eval=FALSE, echo=TRUE}
# Attach the brms package
library(brms)

# Run a Bayesian model
rt_bm <- brm(
  # This is the formula of the model.
  RT ~ 1,
  # This is the probability distribution family.
  family = gaussian(),
  # And the data.
  data = mald
)
```

---

```{r rt-bm-sum}
rt_bm
```

---

```{r rt-bm-sum-2}
summary(rt_bm, prob = 0.7)
```


---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & \sim Gaussian(1046, 6.26) \\
\sigma & \sim Gaussian_+(347, 4.38)
\end{aligned}
$$
]

--

- There is a 95% probability that $\mu$ is between 1034 and 1059 ms. There is a 70% probability that it is between 1040 and 1053.

- There is a 95% probability that $\sigma$ is between 339 and 356 ms. There is a 70% probability that it is between 343 and 352.

--

Great! But what about the RT when the word is a real word vs when it is not?

---

layout: false
layout: true

## Real vs nonce words

---

```{r rt-bm-2, echo=FALSE}
rt_bm_2 <- brm(
  RT ~ IsWord,
  data = mald,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/rt_bm_2",
  seed = my_seed
)
```

```{r rt-bm-2-code, eval=FALSE, echo=TRUE}
# Attach the brms package
library(brms)

# Run a Bayesian model
rt_bm_2 <- brm(
  # This is the formula of the model.
  RT ~ IsWord,
  # This is the probability distribution family.
  family = gaussian(),
  # And the data.
  data = mald
)
```

---

```{r rt-bm-2-sum}
summary(rt_bm_2, prob = 0.89)
```

---

```{r rt-bm-2-sum-2}
cat(capture.output(summary(rt_bm_2, prob = 0.89))[1:15], sep = "\n")
```


.bg-washed-yellow.b--gold.ba.bw2.br3.shadow-5.ph4.mt2[
- The `Intercept` is the mean RT when [IsWord=TRUE].

- `IsWordFALSE` is the **difference** between the mean RT when [IsWord=FALSE] and the mean RT when [IsWord=TRUE].
]

---

```{r rt-bm-2-sum-3}
cat(capture.output(summary(rt_bm_2, prob = 0.89))[1:15], sep = "\n")
```


.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
There is an **89% probability** that the mean RT when [IsWord=FALSE] is **between 113 to 152 ms longer** than the mean RT when [IsWord=TRUE].
]

---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & = \beta_0 + \beta_1 \times IsWord_{F} \\
\end{aligned}
$$
]

.bg-washed-yellow.b--gold.ba.bw2.br3.shadow-5.ph4.mt2[
- $\beta_0$ is mean RT when [IsWord=TRUE].

- $\beta_1$ is the **difference** between the mean RT when [IsWord=FALSE] and the mean RT when [IsWord=TRUE].

- $IsWord_F$ is `0` when [IsWord=TRUE] and `1` when [IsWord=FALSE].
]


---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & = \beta_0 + \beta_1 \times IsWord_{F} \\
\end{aligned}
$$
]

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
**Mean RT when [IsWord=TRUE].**

$$
\begin{aligned}
\mu & = \beta_0 + \beta_1 \times IsWord_{F} = \beta_0 + \beta_1 \times 0 = \beta_0 \\
\end{aligned}
$$
**Mean RT when [IsWord=FALSE].**

$$
\begin{aligned}
\mu & = \beta_0 + \beta_1 \times IsWord_{F} = \beta_0 + \beta_1 \times 1 = \beta_0 + \beta_1 \\
\end{aligned}
$$
]


---

.f3[
$$
\begin{aligned}
RT_i & \sim Gaussian(\mu, \sigma) \\
\\
\mu & = \beta_0 + \beta_1 \times IsWord_{F} \\
\beta_0 & \sim Gaussian(\mu_0, \sigma_0) \\
\beta_1 & \sim Gaussian(\mu_1, \sigma_1) \\
\\
\sigma & \sim Gaussian_+(\mu_2, \sigma_2)
\end{aligned}
$$
]
