---
title: "Bayesian Linear Models"
subtitle: "03 - Interactions"
author: "Stefano Coretta"
institute: "University of Edinburgh"
date: "2023/05/16"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:
      - default
      - "../custom.css"
      - "../xaringan-themer.css"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "../macros.js"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, out.height = "500px", dpi = 300,
  fig.align = "center", fig.width = 7, fig.height = 5
)
knitr::opts_knit$set(root.dir = here::here())

library(tidyverse)
theme_set(theme_minimal())
library(glue)

library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))

options(htmltools.dir.version = FALSE)

library(brms)

my_seed <- 9899
```

```{r mald, echo=FALSE}
mald <- readRDS("./data/mald.rds")
```

---

layout: true

## Many predictors

---

```{r rt-int}
mald %>%
  ggplot(aes(PhonLev, RT)) +
  geom_point(alpha = 0.05) +
  geom_smooth(aes(colour = IsWord, fill = IsWord), method = "lm")
```

---

```{r rt-bm-3, echo=TRUE}
my_seed <- 9899

rt_bm_3 <- brm(
  RT ~ IsWord + PhonLev,
  data = mald,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/rt_bm_3",
  seed = my_seed
)
```

---

```{r rt-bm-3-sum}
cat(capture.output(summary(rt_bm_3, prob = 0.89))[8:12], sep = "\n")
```


.bg-washed-yellow.b--gold.ba.bw2.br3.shadow-5.ph4.mt2[
- The `Intercept` is the mean RT when [IsWord=TRUE] and [PhonLev=0].

- `IsWordFALSE` is the **difference** between the mean RT when [IsWord=FALSE] and the mean RT when [IsWord=TRUE].

- `PhonLev` is the **difference** between the mean RT when [PhonLev=1] and [PhoneLev=0].
]

---

```{r rt-bm-3-plot}
conditional_effects(rt_bm_3, "PhonLev:IsWord")
```

---

```{r rt-bm-3-plot-2}
conditional_effects(rt_bm_3, "IsWord:PhonLev")
```

---

```{r rt-int-2}
mald %>%
  ggplot(aes(PhonLev, RT)) +
  geom_point(alpha = 0.05) +
  geom_smooth(aes(colour = IsWord, fill = IsWord), method = "lm")
```

---

layout: false
layout: true

## We need an interaction

---

```{r rt-bm-4, echo=TRUE}
rt_bm_4 <- brm(
  RT ~ IsWord + PhonLev + IsWord:PhonLev,
  data = mald,
  backend = "cmdstanr",
  cores = 4,
  file = "./data/cache/rt_bm_4",
  seed = my_seed
)
```

.bg-washed-blue.b--dark-blue.ba.bw2.br3.shadow-5.ph4.mt2[
- `isWord:PhonLev` is called an **interaction** term.
- `isWord:PhonLev` allows the model to capture differences in the effect of `isWord` depending on the value of `PhonLev` (and vice versa).
]

---

```{r rt-bm-4-sum}
cat(capture.output(summary(rt_bm_4, prob = 0.89))[8:13], sep = "\n")
```

.bg-washed-yellow.b--gold.ba.bw2.br3.shadow-5.ph4.mt2[
- The `Intercept` is the mean RT when [IsWord=TRUE] and [PhonLev=0].

- `IsWordFALSE` is the **difference** between the mean RT when [IsWord=FALSE] and the mean RT when [IsWord=TRUE], when [PhonLev=0].

- `PhonLev` is the **difference** between the mean RT when [PhonLev=1] and [PhoneLev=0], when [IsWord=TRUE].

- `IsWordFALSE:PhonLev` is the adjustment to the difference in mean RT when [IsWord=FALSE] and [PhonLev=1].
]

---

```{r rt-bm-4-plot}
conditional_effects(rt_bm_4, "PhonLev:IsWord")
```
